Surrogate model of aligned-spin binary neutron star waveforms using Gaussian process regression

### 2018-01-16 paper/paper.tex
Initial surrogate of EOB with quadrupole-monopole effect. There's still some discrepency in the phase that needs to be checked.

\author{
Benjamin D. Lackey$^1$, 
Michael P\"{u}rrer$^1$
Andrea Taracchini$^1$
}
\affiliation{
$^1$Max Planck Institute for Gravitational Physics, Albert Einstein Institute, D-14476 Golm, Germany
}

@@

Fast and accurate waveform models are necessary for measuring properties of inspiraling binary neutron star systems such as the masses, spins, and tidal parameters. The effective one body (EOB) formalism has been used to produce BNS waveforms that include aligned-spin, quadrupolar and octopolar adiabatic tides, as well as the effect of dynamical tides, and the formalism is accurate enough to avoid significant biases in the measured parameters. Unfortunately, current implementations take several minutes to a day to evaluate, and this is too slow for parameter estimation codes that require $10^7$--$10^8$ sequential waveform evaluations. We construct a frequency-domain surrogate of this model that requires less than 0.3~s to evaluate for a waveform with a starting frequency of 10~Hz (0.04~s for a starting frequency of 30~Hz). We first reduce the 12-dimensional intrinsic parameter space of this model to the 5-dimensional parameter space that includes mass ratio, aligned-spins, and quadrupolar tides using universal relations that approximate all matter effects in terms of the leading quadrupolar tidal parameter. We then use Gaussian process regression and active learning to optimally sample the parameter space and construct a surrogate with $\sim 1000$ waveforms instead of the $\gtrsim 10^5$ waveforms that would be required with standard, grid-based interpolation methods. \red{[[Mismatch is ...]]}

@@

The resulting waveforms still have some remaining Gibbs oscillations which can be seen as small-amplitude, high-frequency oscillations in the residuals $\Delta\ln(A)(Mf)$ and $\Delta\Phi(Mf)$. These come from two sources. The first is the fact that the Planck window at the beginning of the waveform was not sufficiently long. We could make this window longer, but that would require us to start the surrogate at a higher frequency. The second source comes from the end of the waveform where the amplitude rapidly drops to zero amplitude after the peak amplitude in $\sim 1$ cycle. The Gibbs oscillations at high frequencies are therefore a genuine feature of the EOB model. In an attempt to preserve all structure in the EOB waveform, we choose not to window the end of the time-domain waveform. We instead remove these oscillations in the frequency domain using a moving average filter centered on $Mf$ with width $[Mf(1-\Delta Mf), Mf(1+\Delta Mf)]$. We use a width of $\Delta Mf=0.1$ for the amplitude residual and $\Delta Mf=0.05$ for the phase residual. Smoothing these oscillations makes it significantly easier to interpolate the amplitude and phase residuals as functions of $Mf$ and $\bx$.

@@

In GPR a function $g(\bx)$ is described in terms of a mean $m(\bx)$ and  covariance $k(\bx, \bx')$ between points.
\begin{equation}
g(\bx) \approx \mathcal{GP}(m(\bx), k(\bx, \bx')).
\end{equation}
The mean $m(\bx)$ can be a parameterized fitting function while $k(\bx, \bx')$ is a kernel with tunable hyperparameters that describes the covariance between the point $\bx$ and the sampled point $\bx'$. This covariance can be used to describe features of the function not captured by the parameterized mean $m(\bx)$ as well as genuine noise in the data $\by$. For the problem here, we have already subtracted the TaylorF2 waveform from EOB waveform, so we set $m(\bx) = 0$ and model the residuals purely in terms of the kernel $k(\bx, \bx')$.

>>>>

\author{
Benjamin D. Lackey, 
Michael P\"{u}rrer, 
Andrea Taracchini,
Sylvain Marsat
}
\affiliation{
Max Planck Institute for Gravitational Physics, Albert Einstein Institute, D-14476 Golm, Germany
}

@@

Fast and accurate waveform models are necessary for measuring the properties of inspiraling binary neutron star systems such as the recently discovered event GW170817. The effective one body (EOB) formalism has been used to produce BNS waveforms that include aligned-spin, quadrupolar and octopolar adiabatic tides, as well as the effect of dynamical tides, and the formalism is accurate enough to avoid significant biases in the measured parameters. Unfortunately, current implementations take several minutes to a day to evaluate, and this is too slow for parameter estimation codes that require $10^7$--$10^8$ sequential waveform evaluations. We construct a frequency-domain surrogate of this model that requires less than 0.3~s to evaluate for a waveform with a starting frequency of 10~Hz (0.04~s for a starting frequency of 30~Hz). We first reduce the 12-dimensional intrinsic parameter space of this model to the 5-dimensional parameter space that includes mass ratio, aligned-spins, and quadrupolar tides using universal relations that approximate all matter effects in terms of the leading quadrupolar tidal parameter. We then use Gaussian process regression and active learning to optimally sample the parameter space and construct a surrogate with $\sim 1000$ waveforms instead of the $\gtrsim 10^5$ waveforms that would be required with standard, grid-based interpolation methods. \red{[[Mismatch is ...]] [[It can be used in arbitrary frequency range by matching to analytic TaylorF2.]] [[Improvement to previously published EOB model [cite] by including spin-induced quadrupole moment and prescription for waveform tapering at merger.]] }

@@

The resulting waveforms still have some remaining Gibbs oscillations which can be seen as small-amplitude, high-frequency oscillations in the residuals $\Delta\ln(A)(Mf)$ and $\Delta\Phi(Mf)$. These come from two sources. The first is the fact that the Planck window at the beginning of the waveform was not sufficiently long. We could make this window longer, but that would require us to start the surrogate at a higher frequency. The second source comes from the end of the waveform where the amplitude rapidly drops to zero amplitude after the peak amplitude in $\sim 1$ cycle. The Gibbs oscillations at high frequencies are therefore a genuine feature of the EOB model. In an attempt to preserve all structure in the EOB waveform, we choose not to window the end of the time-domain waveform. We instead remove these oscillations in the frequency domain using a moving average filter centered on $Mf$ with width $[Mf(1-\Delta Mf), Mf(1+\Delta Mf)]$. We use a width of $\Delta Mf=0.1$ for the amplitude residual and $\Delta Mf=0.05$ for the phase residual. Smoothing these oscillations makes it significantly easier to interpolate the amplitude and phase residuals as functions of $Mf$ and $\bx$. \red{[[$\Delta MF$ is a difference not a fraction. maybe try $p$ or something similar? Mention that 0.1 for the amplitude is a bit aggressive, but only matters for high frequency where the amplitude is crazy anyway.]]}

@@

In GPR a function $g(\bx)$ is described in terms of a mean $m(\bx)$ and  covariance $k(\bx, \bx')$ between points.
\begin{equation}
g(\bx) \approx \mathcal{GP}(m(\bx), k(\bx, \bx')).
\end{equation}
The mean $m(\bx)$ can be a parameterized fitting function while $k(\bx, \bx')$ is a kernel with tunable hyperparameters that describes the covariance between the point $\bx$ and the sampled point $\bx'$. This covariance can be used to describe features of the function not captured by the parameterized mean $m(\bx)$ as well as genuine noise in the data $\by$. For the problem here, we have already subtracted the TaylorF2 waveform from the EOB waveform, so we set $m(\bx) = 0$ and model the residuals purely in terms of the kernel $k(\bx, \bx')$.
